---
title: "Diabetes vs HDL Levels"
format: 
  pdf:
    toc: true
    number-sections: true
    colorlinks: true
editor: visual
---

# example change

# Background

For this study, the NHANES dataset was used to analyze the association between diabetes and high-density lipoprotein (HDL) cholesterol levels. High levels of HDL decrease the risk of cardiovascular disease by transporting excess cholesterol from peripheral tissues to the liver for excretion. However, research has shown that diabetes tends to decrease HDL levels, which is a risk factor for cardiovascular disease [(Femlak et al., 2017)](https://doi.org/10.1186/s12944-017-0594-3). Informing individuals with a diabetic background about this association can help patients be more aware of their body and risk factors throughout their lives.

# Research Question and Hypothesis

Research Question: Is there an association between diabetes and HDL cholesterol levels among adults aged 18 and older?

Null Hypothesis (H0): There is no difference in the average HDL cholesterol levels (mmol/L) between adults (aged 18 and older) with diabetes and those without diabetes.

Alternative Hypothesis (H1): There is a difference in the average HDL cholesterol levels (mmol/L) between adults (aged 18 and older) with diabetes and those without diabetes.

# Data Description and Exploratory Data Analysis

For this research study, the data was obtained from the US National Health and Nutrition Examination Study survey, version 2.1.0.

```{r, message=FALSE, warning=FALSE}
#load library
library(NHANES)
library(dplyr)
library(ggplot2)
head(NHANES)
```

The targeted variable here is the average HDL levels (mmol/L), which is a numerical variable, and it is being compared between two groups: diabetes and non-diabetes (categorical) with. The normal range for HDL levels differs between genders and between adults and children [(Mayo Clinic)](https://www.mayoclinic.org/diseases-conditions/high-blood-cholesterol/in-depth/hdl-cholesterol/art-20046388). To eliminate these impacts, analyzed female and male adults (ages 18 and older) separately.

```{r}
filter_data <- NHANES %>%
  filter(Age >= 18) %>%
  select(Age, Gender, Diabetes, DirectChol) %>%
  na.omit()
summary(filter_data)

# Females with Diabetes
female_diabetes <- NHANES %>%
  filter(Gender == "female", Diabetes == "Yes") %>% 
  select(Age, Gender, Diabetes, DirectChol) %>% 
  na.omit()
summary(female_diabetes)

# Females with no Diabetes
female_no_diabetes <- NHANES %>%
  filter(Gender == "female", Diabetes == "No") %>% 
  select(Age, Gender, Diabetes, DirectChol) %>% 
  na.omit()
summary(female_no_diabetes)

# Male with Diabetes
male_diabetes <- NHANES %>%
  filter(Gender == "male", Diabetes == "Yes") %>% 
  select(Age, Gender, Diabetes, DirectChol) %>% 
  na.omit()
summary(male_diabetes)

# Male with no Diabetes
male_no_diabetes <- NHANES %>%
  filter(Gender == "male", Diabetes == "No") %>% 
  select(Age, Gender, Diabetes, DirectChol) %>% 
  na.omit()
summary(male_no_diabetes)

```

The mean HDL cholesterol values are 1.31 mmol/L and 1.51 mmol/L for the diabetic and non-diabetic groups in females, respectively. For males, the mean values are 1.18 mmol/L and 1.25 mmol/L for the diabetic and non-diabetic groups, respectively. Consequently, the non-diabetic group has a higher mean HDL cholesterol level for both genders.

# Visualize Data

```{r}
# Create box plot 
ggplot(filter_data, aes(x = interaction(Gender, Diabetes), 
                        y = DirectChol, fill = Gender)) + 
  geom_boxplot(alpha = 0.7) + 
  labs(title = "Cholesterol Levels by Gender and Diabetes Status"
       , x = "Group", y = "Direct Cholesterol Levels", fill = "Gender") + theme_minimal()
#creat histogram
ggplot(filter_data, aes(x = DirectChol, fill = interaction(Gender, Diabetes))) +
  geom_histogram(position = "identity", bins = 40, alpha = 0.7) +
  scale_fill_manual(values = c("blue", "red", "green", "purple")) +
  labs(title = "Cholesterol Levels by Gender and Diabetes Status",
       x = "Direct Cholesterol Levels (mmol/L)", y = "Count", fill = "Group") + theme_bw()
```

Box plot shows outliers for all four groups, which may cause a slightly positive skew that is also indicated by the histogram.

# Analysis

```{r}
# ANOVA model
model <- aov(DirectChol ~ Diabetes * Gender + Age, data = filter_data)
summary(model)


```

## **Diabetes:**

**F value**: 85.89

**p-value**: \< 2e-16

**Significance**: Highly significant

**Interpretation**: There is a strong association between diabetes and HDL cholesterol levels. Individuals with diabetes have different HDL cholesterol levels compared to those without diabetes, even after adjusting for gender and age.

## **Gender:**

**F value**: 934.12

**p-value**: \< 2e-16

**Significance**: Highly significant

**Interpretation**: Gender has a highly significant impact on HDL cholesterol levels. Males and females have significantly different HDL cholesterol levels after adjusting for diabetes status and age.

## **Age:**

**F value**: 142.87

**p-value**: \< 2e-16

**Significance**: Highly significant

**Interpretation**: Age significantly affects HDL cholesterol levels. This suggests that as age varies, there is a significant change in HDL cholesterol levels, independent of gender and diabetes status.

## **Interaction between Diabetes and Gender (Diabetes:Gender):**

**F value**: 24.20

**p-value**: 8.89e-07

**Significance**: Highly significant

**Interpretation**: There is a significant interaction between diabetes and gender on HDL cholesterol levels. This indicates that the effect of diabetes on HDL cholesterol levels differs between males and females.

As a result, there is sufficient evidence to conclude that statistically, there is a significant difference in the mean HDL levels between the diabetes and non-diabetes adult women aged 18 and older.

# Conclusions

Lower HDL levels in the bloodstream increase the risk of cardiovascular disease. Based on this study, the HDL levels in people with diagnosed diabetes are lower compared to non-diabetic people, which is consistent with other reports. As a result, **statistically**, people with diabetes are at a higher risk for cardiovascular disease compared to healthy individuals.

According to the Mayo Clinic, adult women are at risk for cardiovascular disease if their HDL cholesterol levels are less than 1.3 mmol/L[(Mayo Clinic)](https://www.mayoclinic.org/diseases-conditions/high-blood-cholesterol/in-depth/hdl-cholesterol/art-20046388). Based on this study, the average HDL level for the diabetic group is 1.31 mmol/L in females, which is close to the risk threshold. This suggests that many women with diabetes might be at borderline risk for cardiovascular disease.

According to the Mayo Clinic, adult men are at risk if their levels are less than 1.0 mmol/L [(Mayo Clinic)](https://www.mayoclinic.org/diseases-conditions/high-blood-cholesterol/in-depth/hdl-cholesterol/art-20046388). Based on this study, the average HDL level for the diabetic group is 1.18 mmol/L in males, which implies that while they might not be at immediate risk, their levels are still close enough to warrant attention.

# References

Femlak, M., Gluba-Brzózka, A., Ciałkowska-Rysz, A., & Rysz, J. (2017). The role and function of HDL in patients with diabetes mellitus and the related cardiovascular risk. *Lipids in Health and Disease, 16*(1), 207. https://doi.org/10.1186/s12944-017-0594-3

Mayo Clinic. (n.d.). *HDL cholesterol: How to boost your ‘good’ cholesterol*. Mayo Clinic. Retrieved September 23, 2024, from <https://www.mayoclinic.org/diseases-conditions/high-blood-cholesterol/in-depth/hdl-cholesterol/art-20046388>

# 
